---
title: "Wright-Fisher Simulation for genetic drift"
author: "Jinliang Yang"
date: "01-27-2022"
output: NULL
---

# Wright-Fisher simulation

Consider a single nucleotide position with two alleles, $A_1$ and $A_2$

In generation $t$ there are $x$ individuals carrying allele $A_1$, which is at frequency $p_t = x/2N$.

This implies that there are $2N - x$ individuals carrying allele $A_2$, which is at frequency $q_t = 1- p_t= 1- x/2N$.

The sampling of alleles for the next generations is equivalent to sampling from a binomial distribution with parameters size = $2N$ and prob= $x/2N$.

Therefore, the mean and variance of $p$ in the next generation for the Wright-Fisher model are:


\begin{align*}
E(p_t) &= p_t \\
Var(p_{t+1}) &= p_tq_t/2N 
\end{align*}


---

# Wright Fisher simulation


```{r}
wright_fisher <- function(N=1000, t=100, A1=100){
  # N: number of diploid individuals. [N=1000, integer]
  # t: number of generations. [t=100, numeric]
  # A1: number of A1 (the minor allele) in the first generation. [A1=10, integer]

  # create a vector 
  x <- 0
  x[1] <- A1

  # Start to loop over m generations
  # binomial sampling at each generation - determines allele count in next generations
  for (i in 2:t){
    # calculate allele freq in the current generation
    k <- (x[i-1])/(2*N)
    # generate a vector with 2N individuals
    n <- seq(0,2*N,1)
    
    # density for the binomial distribution with paramters size=2N and prob=k
    prob <- dbinom(n, 2*N, k)
    # plot(prob)
    # choose one from a vector each with certain prob
    x[i] <- sample(0:(2*N), 1, prob=prob)
  }
  # return back the results
  return(x)
}

# sim1 <- wright_fisher(N=50, t=100, A1=20)
```

---

```{r}
set.seed(1234587)
sim1 <- wright_fisher(N=50, t=1000, A1=20)
plot(sim1[1:100], type="o", pch=19, xlab="Generations", ylab="A1 allele count")

sim1 <- wright_fisher(N=50, t=1000, A1=20)
plot(sim1[1:100], type="o", pch=19, xlab="Generations", ylab="A1 allele count")

```


```{r}
set.seed(123456)
sim1 <- wright_fisher(N=50000, t=100, A1=20000)
#plot(sim1[1:100], type="o", pch=19, xlab="Generations", ylab="A1 allele count")
write.table(sim1, "cache/sim1_hcc.txt", sep="\t", row.names=FALSE, quote=FALSE)


#
sim1 <- read.table("../../cache/sim1_hcc.txt", header=T)
plot(sim1$x, type="o", pch=19, xlab="Generations", ylab="A1 allele count")
```


